---
title: "Categorical Predictors & Interaction terms"
subtitle: "IDS 702"
format: html
---

**Load packages and data**

```{r load-libraroies, warning = F, message = F}
library(tidyverse)
library(tidymodels)
library(palmerpenguins)

```

Today's data is called "penguins" from the palmerpenguins library. The data were collected and made available by [Dr. Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php) and the [Palmer Station, Antarctica LTER](https://pallter.marine.rutgers.edu/), a member of the [Long Term Ecological Research Network](https://lternet.edu/).

## In your group, address the following:

-   What are the sample size and number of variables?

-   Which variables are categorical and which are numeric? Are the categorical variables stored appropriately?

-   Is there any missing data? If so, create a new data frame to filter out the missing observations. 11 observations wi

```{r}
data <- penguins

colSums(is.na(data))

data_clean <- na.omit(data)
```

Next, create three plots:

-   a scatter plot to illustrate the relationship between bill length and body mass. Describe what you see in the plot.

-   a scatter plot of bill length and body mass, with the points colored by sex. You can use the code below as a template. Describe what you see.

-   a scatter plot of bill depth and body mass, with the points colored by species. Describe what you see.

```{r}
ggplot(data, aes(x=bill_length_mm, y=body_mass_g))+
  geom_point()+
  geom_smooth(method="lm",se=F)
```

so here we can see that there is positive relationship between bill length and body mass. Linear model seems reasonable here.

```{r, eval=FALSE}
ggplot(data_clean, aes(x=bill_length_mm, y=body_mass_g, color=sex
                 ))+
  geom_point()+
  geom_smooth(method="lm",se=F)
```

we want to see if there is some interaction term between bill length and body mass, here we can see the variable sex is not affected as an interaction term. Males have a higher body mass.

```{r, eval=FALSE}
ggplot(data_clean
       , aes(x=bill_depth_mm, y=body_mass_g, color=species
                 ))+
  geom_point()+
  geom_smooth(method="lm",se=F) #line of best fit for a data
```

we want to see if there is some interaction term between bill length and body mass, here we can see the variable species do have an affected as an interaction term. Adelie and Chinstap have the same body mass, but the Gentoo is different.

## Categorical predictors and interaction terms

**Categorical Predictors**

We will use **dummy variables** to represent categorical factors in the regression model. These variables take a value of 1 for a certain level, and 0 otherwise. We always have \# levels -1 dummy variables (**why?**)

When we have a categorical predictor in the model, we are fitting **parallel lines**. In this case, we are changing the **intercept** based on the levels of the categorical predictor.

Let's say we regress body mass on bill length and sex. Mathematically, the model would be:

$$
Y = \beta_0 + \beta_1x_1 + \beta_2x_2 + \epsilon
$$

where $x_1$ = bill length (mm), and $x_2$ = 1 when sex=male, and 0 otherwise.

Thus, for male penguins, the model is: $Y=(\beta_0+\beta_2)+\beta_1x_1+\epsilon$, and for female penguins, the model is: $Y=\beta_0+\beta_1x_1+\epsilon$

Let's fit the model:

```{r}
mod1_sex <- lm(body_mass_g ~ bill_length_mm + sex, data=data)

summary(mod1_sex)
```

Fitted models: for male penguins $Y=(746.063+405.243)+74.025x_1$

for females penguins \$\$Y=746.063 + 74.025x_1\$\$

Interpretations: controlling for sex per mm increase in bill length body mass in g increases by 74 on average. This relationship is statistically significant with p value of 2e-16.

male penguins has 405 more mass than females on average. This difference is statistically significant with the p value of 5.43e-08

**Interaction terms**

When we have an interaction term in the model, we are fitting lines with **different slopes**. Interaction terms allow us to assess how the **relationship between a predictor and an outcome changes based on the value of another predictor**. We will focus on interaction terms between a continuous predictor and a categorical predictor.

Let's regress body mass on bill length with an interaction term with species. Mathematically, the model would be:

$$
Y = \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_4x_1x_2 + \beta_5x_1x_3 + \epsilon
$$

where $x_1$ is bill length, $x_2$=1 for Chinstrap penguins, 0 otherwise, and $x_3$=1 for Gentoo penguins, 0 otherwise

Then, we can write species-specific models.

Adelie penguins: $Y=\beta_0+\beta_1x_1+\epsilon$

Chinstrap penguins: $Y=(\beta_0+\beta_2) + (\beta_1+\beta_4)x_1+\epsilon$

Gentoo penguins: $Y=(\beta_0+\beta_3)+(\beta_1+\beta_5)x_1+\epsilon$

Let's fit the model:

```{r}
mod2_species <- lm(body_mass_g~bill_length_mm*species, data=data_clean)
summary(mod2_species)
```

Fitted models:

-   Adelie penguins: $\hat{bodymass} = 66.5 + 93.7(bill length)$

-   Chinstrap penguins: $\hat{bodymass} = (66.5 + 779.69) + (93.7 - 34.63)(billlength)$

-   Gentoo: $bodymass = (66.5 -94.35) + (93.7 + 13.69)(billength)$

Interpretations:

-   \$ beta 1 hat - For Adelie penguins, per mm increase in bill length, body mass increases by 93.7g on average.

-   \$\hat\beta\_4\$ - For Chinstrp penguins, per mm increase in bill length, body mass decreases by 34.6 g compared to Adelie penguins.

-   for Gentoo penguins, per mm increase in bill length, body mass increases by 107.7 (93.75 + 13.89) gram on average.

**Nested F test**

We often want to know if a categorical predictor or an interaction term **as a whole** is statistically significant. Because we have multiple coefficients for these terms, we need an alternative approach to assess the factors as a whole. We can do this with nested F tests, which compare a **full model** to a **reduced model**.

Let's assess whether or not the species interaction term is statistically significant as a whole:

The full model is what we fit above: $Y = \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_4x_1x_2 + \beta_5x_1x_3 + \epsilon$

The reduced model is the model without the interaction term: $Y = \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon$

This test will compare the sum of squared errors for the two models. Remember that a lower SSE indicates a better model. If the SSE is significantly lower for the full model than the reduced model, the full model would be considered a better model, and we would say the term being test is statistically significant.

We can use the `anova` function to conduct the test:

```{r}
mod_full <- mod2_species
mod_reduced <- lm(body_mass_g~bill_length_mm+species, data=data_clean)

anova(mod_reduced, mod_full, test="F")
```

We do not need to do this if the categorical variable only has two levels. **Why?**

Why? we will have one term in the model, we c

```{r}
plot(mod2_species)

```

an only look at its p value instead of ....

neither of this are statistically significant. different test that we are running.

## Exercise

1.  Create a scatter plot of bill depth and body mass. Color the plot by sex and then species. For which of these variables, if either, does an interaction term appear to be useful? Describe your reasoning.

2.  Fit a model regressing body mass on bill depth and species without an interaction term. Write the fitted models for each species. Write interpretations for the coefficient estimates.

3.  Fit a model regressing body mass on bill depth and species with an interaction term. Write the fitted models for each species. Write interpretations for the interaction coefficient estimates.

4.  Conduct a nested F test to assess the statistical significance of the species interaction term as a whole.

## 
