---
title: "add_20241214"
author: "Eleanor Jiang 305002785"
date: "2024-12-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(knitr)
```

```{r preprocessing, warning=FALSE}
data <- read.csv("police_killings.csv")
# Data cleaning: filter missing values and process variables
cleaned_data <- data %>%
  mutate(
    age = as.numeric(age),
    urate = as.numeric(urate),
    pov = as.numeric(pov),
    share_white = as.numeric(share_white),
    share_black = as.numeric(share_black),
    share_hispanic = as.numeric(share_hispanic),
    p_income = as.numeric(p_income),
    gender = as.factor(gender),
    raceethnicity = ifelse(raceethnicity == "Unknown", NA,raceethnicity)
  ) %>%      
  filter(!is.na(raceethnicity) & !is.na(comp_income) & !is.na(pov) & !is.na(age) & !is.na(urate) & !is.na(armed)) %>%
  mutate(
    armed = factor(ifelse(armed == "No", "No", "Yes")),
      raceethnicity = factor(raceethnicity),  
         raceethnicity = relevel(raceethnicity, ref = "White")  
  )

# Display summary of cleaned data
glimpse(cleaned_data)
```

## EDA

```{r eda1}
par(mfrow = c(2, 1))
hist(cleaned_data$age, breaks = 20, main = "Age Distribution", xlab = "Age")
hist(cleaned_data$urate, breaks = 20, main = "Unemployment Rate Distribution", xlab = "Unemployment Rate")
```

The demographic analysis shows that the deceased had a mean age in their 30s, with the age range extending to over 80. The unemployment rate in their residential tracts averaged 10%, reaching a maximum of 51%. The gender distribution was heavily skewed, with 428 males and 20 females.


```{r, eda2, message=FALSE, fig.align='center'}
race_summary <- cleaned_data %>%
  count(raceethnicity) %>%
  mutate(prop = n / sum(n)) %>%
  kable()
rq1_data <- cleaned_data %>% filter(raceethnicity %in% c("Black", "White", "Hispanic/Latino"))

library(reshape2)
library(ggplot2)
library(viridis)

options(repr.plot.width = 5, repr.plot.height = 1) 
ggplot(rq1_data, aes(x = comp_income, y = raceethnicity, fill = raceethnicity)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_viridis(discrete = TRUE) +
  labs(title = "Economic Conditions and Racial Composition",
       x = "Relative Income",
       y = "Race/Ethnicity") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "none"
  ) +
  coord_flip()

# Create correlation matrix
cor_matrix <- cor(rq1_data[, c("comp_income", "pov", "nat_bucket", "county_bucket", "p_income", "urate")], use = "complete.obs")

# Melt the correlation matrix
cor_melted <- melt(cor_matrix)

name_mapping <- c(
  "comp_income" = "Relative Income",
  "pov" = "Poverty Rate",
  "nat_bucket" = "National Income Tier",
  "county_bucket" = "County Income Tier",
  "p_income" = "Personal Income Tier",
  "urate" = "Unemployment Rate"
)

library(scales)
options(repr.plot.width = 5, repr.plot.height =4) 
# Create correlation plot with viridis colors
ggplot(cor_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_viridis(option = "viridis", limits = c(-1, 1)) +
  geom_text(aes(label = round(value, 2)), 
            color = ifelse(abs(cor_melted$value) > 0.5, "white", "black"), 
            size = 4) +
  labs(title = "Correlation Plot of Economic Related Indicators",
       x = "", y = "") +
  scale_x_discrete(labels = \(x) label_wrap(10)(name_mapping[x])) +
  scale_y_discrete(labels = \(x) label_wrap(10)(name_mapping[x])) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 10),
   # axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_blank(),
    panel.border = element_blank()
  )
```
Due to sample size limitations, the racial composition analysis was restricted to three main groups: White (52%), Black, and Hispanic/Latino, with Asian and Native American groups (each <5%) with less than 30 people being excluded from the analysis. For our **research question 1**, it is clear that a multinomial logistic regression is proficient in this case. The box plot reveals distinct income disparities across racial groups, suggesting relative income as a potentially significant predictor. Analysis of economic indicators showed high correlations among most variables, with tract-level unemployment rate being the exception. Based on this correlation analysis, we selected unemployment rate and one additional economic indicator for the final model.
