---
title: "eda"
author: "Eleanor Jiang 305002785"
date: "2024-11-17"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## EDA


```{r basic EDA, echo=FALSE}
library(dplyr)
library(ggplot2)

# Read the new CSV file
new_data <- read.csv("police_killings_processed.csv")

## Examine NAs
print(sapply(new_data, function(x) sum(is.na(x))))

## Convert age to numeric

# Convert the 'age' column to numeric
new_data$age_num <- as.numeric(new_data$age)

# Verify the conversion
class(new_data$age_num)

# Check for any NAs introduced during conversion
sum(is.na(new_data$age_num))

# Check the original data
# print(new_data[is.na(new_data$age_num), c("age", "age_num")])

# Convert "Unknown" age to mean age
new_data$age <- as.numeric(ifelse(new_data$age == "unknown", NA, new_data$age))
new_data$age[is.na(new_data$age)] <- mean(new_data$age, na.rm = TRUE)
```

The indicator `age` contains missing values "Unknown". We choose to fill them with the average value.


```{r}
glimpse(new_data)


####### Research question 1
## 1. Race/Ethnicity
race_counts <- table(new_data$raceethnicity)
race_percentages <- prop.table(race_counts) * 100

cat("Race/Ethnicity Distribution:\n")
print(race_counts)
cat("\nPercentages:\n")
print(race_percentages)

plot1 <- ggplot(new_data, aes(x = raceethnicity)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of Race/Ethnicity", x = "Race/Ethnicity", y = "Count")

# Plot 1: Relationship between 'race' and 'poverty rate'
plot2 <- ggplot(new_data, aes(x = raceethnicity, y = pov)) +
  geom_boxplot(fill = "lightgreen", outlier.colour = "red", outlier.shape = 1) +
  geom_jitter(width = 0.2, alpha = 0.2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Relationship between Race/Ethnicity and Poverty Rate",
       x = "Race/Ethnicity",
       y = "Poverty Rate")

#### Research question 2

## 6. Armed Status
armed_counts <- table(new_data$armed)
armed_percentages <- prop.table(armed_counts) * 100

cat("\nArmed Status Distribution:\n")
print(armed_counts)
cat("\nPercentages:\n")
print(armed_percentages)

plot3 <- ggplot(new_data, aes(x = armed)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of Armed Status", x = "Armed Status", y = "Count")

# Plot 2: Relationship between 'armed' and 'age'
plot4 <- ggplot(new_data, aes(x = armed, y = age)) +
  geom_boxplot(fill = "lightblue", outlier.colour = "red", outlier.shape = 1) +
  geom_jitter(width = 0.2, alpha = 0.2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Relationship between Armed Status and Age",
       x = "Armed Status",
       y = "Age")

library(cowplot)
plot_grid(plot1, plot2, plot3, plot4, ncol=2)

```

There are mostly black and white in race/ethnicity in police killings. We can see that the average poverty rate is lower in white compared to black. Moreover, we can observe that there are many people armed with firearm and the range of the age of the deceased is wide from less than 30 to 90.
