---
title: "IDS 702 Final Report: Analysis of Police-Related Incidents: Economic and Demographic Influences"
author: "Afag, Peter, Eleanor, Mobasser"
date: "`r Sys.Date()`"
format: pdf
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
data <- read.csv("data/police_killings.csv")
library(tidyverse)
library(knitr)
library(pander)
library(caret)
knitr::opts_chunk$set(echo = FALSE)
```

## Abstract

This report investigates two key research questions about police-related incidents: (1) whether there is an association between the economic conditions of a county and the racial composition of individuals involved, and (2) whether the likelihood of an individual being armed during such incidents varies based on their age and the unemployment rate (`urate`) in the area. Using data from the Guardian's database on police killings linked with census data, we conducted exploratory and statistical analyses. Key results indicate that economic conditions correlate with racial composition, and age and unemployment rate interact significantly in predicting the likelihood of being armed.

## Introduction

Police-related incidents have become a focal point of public safety research and policy discussions in recent years. The complex interplay between socioeconomic factors, racial dynamics, and law enforcement practices has sparked intense debate and calls for reform across the United States. Understanding the underlying patterns and potential contributing factors to these incidents is crucial for developing effective strategies to enhance community safety and promote equitable policing practices.

Research has shown that economic conditions and racial demographics can significantly influence crime rates and police-community interactions. Studies have found that neighborhoods with higher concentrations of low-income residents and residents of color tend to experience higher rates of police-related deaths. Additionally, unemployment and economic stress have been associated with increased rates of certain types of crime, potentially affecting police encounters. However, the relationship between economic factors and crime is complex, with some research suggesting that economic downturns may actually decrease criminal opportunities in certain contexts.

This report aims to contribute to this body of knowledge by examining two key research questions:

1. Is there an association between the economic conditions of a county and the racial composition of individuals involved in police-related incidents?

2. Does the likelihood of an individual being armed during such incidents vary based on their age and the unemployment rate in their area?

By analyzing data from the Guardian's database on police killings, linked with the American Community Survey (2015), this study seeks to provide insights into these critical questions (Guardian, 2015). The dataset, which includes demographic, economic, and geographic information, offers a unique opportunity to explore the potential relationships between socioeconomic factors and the characteristics of police-related incidents. Understanding these dynamics can inform evidence-based policy-making and interventions aimed at improving community safety and promoting equity in law enforcement practices.

## Methods

```{r clean-data, include=FALSE}
# Data cleaning: filter missing values and process variables
print('here')
cleaned_data <- data %>%
  mutate(
    age = as.numeric(age),
    urate = as.numeric(urate),
    pov = as.numeric(pov),
    share_white = as.numeric(share_white),
    share_black = as.numeric(share_black),
    share_hispanic = as.numeric(share_hispanic),
    p_income = as.numeric(p_income),
    raceethnicity = ifelse(raceethnicity == "Unknown", NA,raceethnicity)
  ) %>%      
  filter(!is.na(raceethnicity) & !is.na(comp_income) & !is.na(pov) & !is.na(age) & !is.na(urate) & !is.na(armed)) %>%
  mutate(
    armed = factor(ifelse(armed == "No", "No", "Yes")),
      raceethnicity = factor(raceethnicity),  
         raceethnicity = relevel(raceethnicity, ref = "White")  
  )

# Display summary of cleaned data
glimpse(cleaned_data)
```

### Data

We used the "police_killings.csv" dataset, which includes variables such as age, racial composition, economic indicators, and whether the individual was armed. Data cleaning involved: - Filtering out missing values for relevant variables (removed `r nrow(data) - nrow(cleaned_data)` rows). - Converting "armed" into a binary factor (armed vs. unarmed) and racial categories into nominal factors. - Ensuring numeric data types for continuous variables like age, `urate`, `comp_income`, and `pov`.

**Edit**:

The dataset consists of **467 observations** and **34 columns**, providing detailed information on police-related incidents across different counties in the United States. The dataset includes variables that capture demographic information (e.g., race, gender, age), geographic details (e.g., city, state, latitude, longitude), economic indicators (e.g., household income, poverty rate, unemployment rate), and other socioeconomic factors. The primary outcome variable, `raceethnicity`, categorizes individuals involved in police-related incidents into racial groups, including "White," "Black," "Hispanic/Latino," "Asian/Pacific Islander," and "Native American." For this analysis, categories with very low representation, such as "Asian/Pacific Islander" and "Native American," were excluded to ensure model reliability.

Key economic variables include: Tract-level median household income as a share of county-level median household income (will be refered as a relative income), poverty rate at the county level, unemployment rate at the county level, the proportion of the population with a Bachelor's degree or higher. The dataset also contains demographic proportions, such as `share_white`, `share_black`, and `share_hispanic`, representing the racial composition within the county.

To prepare the data for analysis, entries with missing values were excluded. Since the number of missing values was minimal (approximately 2 or 3 for some variables), these rows were dropped without significantly affecting the overall dataset size or quality. Additionally, the outcome variable was structured to focus on three primary racial groups: "White," "Black," and "Hispanic/Latino," with "White" designated as the reference category. This decision was made due to the disproportionately low representation of individuals categorized as "Asian/Pacific Islander" and "Native American," which would have hindered the reliability of the multinomial logistic regression analysis.

### Model fitting and evaluation

The models were selected based on the specific research questions and the nature of the variables involved. For Research Question 1, which examines the relationship between economic conditions and racial composition, a multinomial logistic regression model was appropriate given the categorical nature of the outcome variable, race/ethnicity, with multiple nominal categories. The independent variables, relative income, poverty level, college level and unemployment rate, were chosen to capture key socioeconomic dimensions influencing racial or ethnic group differences. This model enables the examination of how economic factors are associated with the likelihood of belonging to different racial or ethnic categories.

For Research Question 2, a logistic regression model was employed to analyze the binary outcome variable, armed (armed status of the individual). The independent variables include age and urate (unemployment rate), which serve as proxies for demographic and socioeconomic factors. An interaction term (age:urate) was included to explore whether the relationship between unemployment and armed status depends on age, allowing for a more nuanced understanding of their combined effects.

Both models were supported by diagnostic evaluations to ensure reliability. Correlation matrix and Variance Inflation Factor (VIF) was used to check for multicollinearity and correlation, and Cook’s distance assessed the influence of individual data points. Classification performance was further evaluated using a confusion matrix, accuracy, precision, recall, and F1 scores, providing a detailed breakdown of the model's predictive capabilities for each category. The area under the Receiver Operating Characteristic (ROC) curve (AUC) was used as an additional metric to assess the model's ability to distinguish between categories. Residual diagnostics, including residual vs. fitted plots, were used to evaluate linearity and homoscedasticity. These diagnostic evaluations and performance metrics ensure that the models are well-suited to address the research questions, effectively capturing the underlying dynamics in the data. All statistical analyses were conducted using R programming language

#### Research Question 1: Economic Conditions and Racial Composition

-   **Outcome Variable**: `raceethnicity` (Nominal) – categorizes the racial/ethnic group of the deceased.
-   **Independent Variables**:
    -   `comp_income`: A measure of relative income (household income divided by county income), indicating economic status.
    -   `pov`: Poverty level in the area, providing additional context on socio-economic conditions.

A multinomial logistic regression model was fitted to examine the association:

-   `raceethnicity ~ comp_income + pov`

#### Research Question 2: Likelihood of Being Armed

-   **Outcome Variable**: `armed` (Binary) – indicates whether the deceased was armed at the time of the incident.
-   **Independent Variables**:
    -   `age`: Age of the deceased (continuous variable).
    -   `urate`: Unemployment rate in the area, representing socio-economic stress.
-   **Interaction Term**: An interaction between `age` and `urate` to assess their combined influence on the likelihood of individuals being armed.

A logistic regression model was fitted:

-   `armed ~ age + urate + age:urate`

### Model Assessment

Model diagnostics included checks for multicollinearity using Variance Inflation Factor (VIF) and assessment of influential points via Cook's distance. For the multinomial model, pseudo-R² and AIC measures were calculated. For the logistic model, residual patterns were assessed for misfit.

## Results

### Exploratory Data Analysis
```{r eda1}
par(mar = c(4, 4, 2, 2)) # Reduce margins
par(mfrow = c(2, 1))
hist(cleaned_data$age, breaks = 20, main = "Age Distribution", xlab = "Age")
hist(cleaned_data$urate, breaks = 20, main = "Unemployment Rate Distribution", xlab = "Unemployment Rate")
```

The demographic analysis shows that the deceased had a mean age in their 30s, with the age range extending to over 80. The unemployment rate in their residential tracts averaged 10%, reaching a maximum of 51%. The gender distribution was heavily skewed, with 428 males and 20 females.

```{r race-summary}
race_summary <- cleaned_data %>% count(raceethnicity) %>% mutate(Percentage = round(n / sum(n) *100,2)) 
kable(race_summary, caption = "Race Ethinicity Summary Statistics")
```

```{r eda2}
rq1_data <- cleaned_data %>% filter(raceethnicity %in% c("Black", "White", "Hispanic/Latino"))

library(reshape2)
library(ggplot2)
library(viridis)

# options(repr.plot.width = 5, repr.plot.height = 1) 
ggplot(rq1_data, aes(x = comp_income, y = raceethnicity, fill = raceethnicity)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_viridis(discrete = TRUE) +
  labs(title = "Economic Conditions and Racial Composition",
       x = "Relative Income",
       y = "Race/Ethnicity") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "none"
  ) +
  coord_flip()

# Create correlation matrix
cor_matrix <- cor(rq1_data[, c("comp_income", "pov", "nat_bucket", "county_bucket", "p_income", "urate")], use = "complete.obs")

# Melt the correlation matrix
cor_melted <- melt(cor_matrix)

name_mapping <- c(
  "comp_income" = "Relative Income",
  "pov" = "Poverty Rate",
  "nat_bucket" = "National Income Tier",
  "county_bucket" = "County Income Tier",
  "p_income" = "Personal Income Tier",
  "urate" = "Unemployment Rate"
)

library(scales)
# options(repr.plot.width = 5, repr.plot.height =4) 
# Create correlation plot with viridis colors
ggplot(cor_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_viridis(option = "viridis", limits = c(-1, 1)) +
  geom_text(aes(label = round(value, 2)), 
            color = ifelse(abs(cor_melted$value) > 0.5, "white", "black"), 
            size = 4) +
  labs(title = "Correlation Plot of Economic Related Indicators",
       x = "", y = "") +
  scale_x_discrete(labels = \(x) label_wrap(10)(name_mapping[x])) +
  scale_y_discrete(labels = \(x) label_wrap(10)(name_mapping[x])) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size =7),
   # axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_blank(),
    panel.border = element_blank()
  )
```

Due to sample size limitations, the racial composition analysis was restricted to three main groups: White (52%), Black, and Hispanic/Latino, with Asian and Native American groups (each <5%) with less than 30 people being excluded from the analysis. For our **research question 1**, it is clear that a multinomial logistic regression is proficient in this case. The box plot reveals distinct income disparities across racial groups, suggesting relative income as a potentially significant predictor. Analysis of economic indicators showed high correlations among most variables, with tract-level unemployment rate being the exception. Based on this correlation analysis, we selected unemployment rate and one additional economic indicator for the final model.


### Research Question 1: Economic Conditions and Racial Composition

The results of the final multinomial logistic regression model are summarized in the table below. The predictors included in the model were chosen based on their relevance to the research question.

The findings reveal important relationships between economic conditions and racial classifications in police-related incidents. Counties with lower **relative income** (a measure of household income as a share of the county median) were more likely to see individuals involved in these incidents categorized as "Black" rather than "White." Similarly, counties with higher **unemployment rates** were associated with an increased likelihood of individuals being classified as "Black" compared to "White." These findings highlight the disproportionate impact of economic inequality and joblessness on Black individuals in the context of police-related incidents.

For "Hispanic/Latino" individuals, the relationships with economic variables were less consistent. For example, relative income showed no significant relationship with the likelihood of individuals being classified as "Hispanic/Latino" versus "White." However, counties with slightly higher **poverty rates** were marginally more likely to have individuals categorized as "Hispanic/Latino" rather than "White," though the effect was small. Counties with a higher proportion of **college-educated residents** were associated with a reduced likelihood of individuals being classified as "Hispanic/Latino" compared to "White," suggesting the potential role of education in shaping these patterns.

The confidence intervals provide a range of values within which the true effects of these factors are likely to fall, and they support the reliability of these findings. Additionally, odds ratios greater than 1 suggest an increase in likelihood, while those less than 1 suggest a decrease. For instance, the odds of being classified as Black increase significantly with higher unemployment, whereas the odds decrease with higher relative income. These conclusions are further supported by the low p-values observed for several predictors in the model, indicating that these relationships are statistically significant.

Overall, these results demonstrate how structural factors like economic inequality and unemployment disproportionately impact certain racial groups in police-related incidents. By identifying these relationships, policymakers and researchers can better understand and address the systemic factors contributing to these disparities.

```{r corr}
cols_to_numeric <- c(
                     "p_income", "h_income", "comp_income", "county_income", 
                     "urate", "college", "pov")
cor_matrix <- cor(cleaned_data[cols_to_numeric])
print(cor_matrix)
```

```{r Excluding the  2 groups}
cleaned_data <- cleaned_data[!(cleaned_data$raceethnicity %in% c("Asian/Pacific Islander", "Native American")), ]
cleaned_data$raceethnicity <- factor(cleaned_data$raceethnicity)
cleaned_data$raceethnicity <- relevel(cleaned_data$raceethnicity, ref = "White")

```

Fitting model on comp_income, poverty rate, unemployment and college, share_black

```{r rq1-model, include=FALSE}
# Multinomial logistic regression
rq1_model2 <- nnet::multinom(
  raceethnicity ~ comp_income + pov + urate + college ,
  data = cleaned_data
)

```

```{r confusion-matrix, include=FALSE}
confusionMatrix(predict(rq1_model2), cleaned_data$raceethnicity, mode="everything")

```

Accuracy of model is still: 58.29

Results:

```{r intercept 2}
intercept_tablee <- as.data.frame(cbind(
  Coefficients = summary(rq1_model2)$coefficients[, "(Intercept)"],
  Std.Errors = summary(rq1_model2)$standard.errors[, "(Intercept)"],
  p.values = 2 * (1 - pnorm(abs(summary(rq1_model2)$coefficients[, "(Intercept)"] /summary(rq1_model2)$standard.errors[, "(Intercept)"])))))
intercept_tablee$Variable <- "Intercept"
intercept_tablee
```

```{r income 2}
comp_income_tablee <- as.data.frame(cbind(
  Coefficients = summary(rq1_model2)$coefficients[, "comp_income"],
  Std.Errors = summary(rq1_model2)$standard.errors[, "comp_income"],
  p.values = 2 * (1 - pnorm(abs(summary(rq1_model2)$coefficients[, "comp_income"] /summary(rq1_model2)$standard.errors[, "comp_income"])))))
comp_income_tablee$Variable <- "Relative Income"
```

```{r poverty 2}
pov_tablee <- as.data.frame(cbind(
  Coefficients = summary(rq1_model2)$coefficients[, "pov"],
  Std.Errors = summary(rq1_model2)$standard.errors[, "pov"],
  p.values = 2 * (1 - pnorm(abs(summary(rq1_model2)$coefficients[, "pov"] /summary(rq1_model2)$standard.errors[, "pov"])))))
pov_tablee$Variable <- "Poverty Rate"

```

```{r urate 2}
urate_tablee <- as.data.frame(cbind(
  Coefficients = summary(rq1_model2)$coefficients[, "urate"],
  Std.Errors = summary(rq1_model2)$standard.errors[, "urate"],
  p.values = 2 * (1 - pnorm(abs(summary(rq1_model2)$coefficients[, "urate"] /
                                 summary(rq1_model2)$standard.errors[, "urate"])))))
urate_tablee$Variable <- "Unemployment Rate"
```

```{r urate}  urate + college + share_black}
college_tablee <- as.data.frame(cbind(
  Coefficients = summary(rq1_model2)$coefficients[, "college"],
  Std.Errors = summary(rq1_model2)$standard.errors[, "college"],
  p.values = 2 * (1 - pnorm(abs(summary(rq1_model2)$coefficients[, "college"] /summary(rq1_model2)$standard.errors[, "college"])))))
college_tablee$Variable <- "Higher Education Share"

```

```{r}
exp(confint(rq1_model2))


```

```{r}

combined_table <- rbind(intercept_tablee, comp_income_tablee, pov_tablee, urate_tablee, college_tablee #, share_black_tablee
                       )

combined_table <- combined_table %>%
  mutate(
    Odds_Ratio = exp(Coefficients),  
    CI_Lower = exp(Coefficients - 1.96 * Std.Errors),  
    CI_Upper = exp(Coefficients + 1.96 * Std.Errors)   
  )
combined_table <- cbind(`Race Ethnicity` = rownames(combined_table), combined_table)


combined_table <- combined_table[, c("Variable", "Coefficients", "Std.Errors", "Odds_Ratio", "CI_Lower", "CI_Upper", "p.values")]


combined_table <- combined_table %>%
  mutate(
    Coefficients = round(Coefficients, 2),
    Std.Errors = round(Std.Errors, 2),
    Odds_Ratio = round(Odds_Ratio, 2),
    CI_Lower = round(CI_Lower, 2),
    CI_Upper = round(CI_Upper, 2),
    p.values = round(p.values, 2)
  )

combined_table$Significance <- ifelse(
  combined_table$p.values < 0.001, "***",
  ifelse(combined_table$p.values < 0.01, "**",
         ifelse(combined_table$p.values < 0.05, "*", ""))
)

kable(combined_table, caption = "Combined Table: Coefficients, Std.Errors, Odds Ratios, Confidence Intervals, and p-values",
      col.names = c("Variable", "Coef.", "SE", "OR", "CI Low", "CI Up", "p-val", "Sig."),
      align = c('l', rep('r', 7)),  # Left align text, right align numbers
      format = "latex",
      booktabs = TRUE)

```


### Research Question 2: Likelihood of Being Armed

The logistic regression model investigates how age, unemployment rate, and their interaction impact the likelihood of being armed. The results indicate no statistically significant relationship between age and armed status (p = 0.7794). This suggests that age alone does not serve as a direct predictor of whether an individual is armed. Similarly, the interaction effect between age and unemployment rate is not significant (p = 0.3639), implying that the combined influence of these factors does not meaningfully affect the likelihood of being armed. However, unemployment rate shows a positive, albeit not statistically significant, relationship with being armed (p = 0.2748). This points to the possibility that higher unemployment rates might indirectly contribute to armed status, potentially through heightened stress or shifts in social dynamics. Although the statistical evidence is not strong, these relationships warrant closer examination in future studies.

```{r rq2-model}
# Logistic regression
rq2_model <- glm(
  armed ~ age + urate + age:urate,
  data = cleaned_data,
  family = binomial
)
pander(summary(rq2_model))

par(mfrow = c(2, 2)) # Set plotting area to 2x2

# Residual vs Fitted
plot(rq2_model$fitted.values, residuals(rq2_model, type = "deviance"),
     xlab = "Fitted Values",
     ylab = "Deviance Residuals",
     main = "Residuals vs Fitted",
     pch = 20, col = "blue")
abline(h = 0, col = "red", lty = 2)

# Normal Q-Q Plot
qqnorm(residuals(rq2_model, type = "deviance"), 
       main = "Normal Q-Q Plot of Residuals",
       pch = 20, col = "blue")
qqline(residuals(rq2_model, type = "deviance"), col = "red", lty = 2)

# Scale-Location Plot
plot(rq2_model$fitted.values, sqrt(abs(residuals(rq2_model, type = "deviance"))),
     xlab = "Fitted Values",
     ylab = "Sqrt |Deviance Residuals|",
     main = "Scale-Location Plot",
     pch = 20, col = "blue")
abline(h = 0, col = "red", lty = 2)

# Cook's Distance
plot(cooks.distance(rq2_model), 
     xlab = "Index",
     ylab = "Cook's Distance",
     main = "Cook's Distance",
     pch = 20, col = "blue")
abline(h = 4/(nrow(cleaned_data) - length(rq2_model$coefficients)), col = "red", lty = 2)

par(mfrow = c(1, 1)) # Reset plotting area
```

```{r resutlss, include=FALSE}
rq2_model_pred <- ifelse(predict(rq2_model, type="response")>0.5,1,0)
rq2_model_pred_fac <- factor(rq2_model_pred,
                            levels=c(0,1),
                            labels=c("No", "Yes"))
confusionMatrix(table(rq2_model_pred_fac, cleaned_data$armed),
                positive="Yes",
                mode="everything")


```

The diagnostic plots suggest areas of concern regarding the model fit. The Residuals vs. Fitted plot reveals non-random patterns, indicating possible model misfit or missing predictors. The Normal Q-Q Plot shows deviations from normality in the residuals, particularly in the tails, which could affect the interpretation of the model. The Scale-Location Plot suggests mild heteroscedasticity, with residual variance increasing at higher fitted values. The Cook’s Distance plot identifies a few influential points, though most observations fall within acceptable limits. These findings indicate the need for further refinement, such as adding predictors or adjusting the model specification, to better capture the data's underlying relationships.

### Distribution of Armed Status

To better understand the prevalence of being armed, Table 7 and the accompanying bar chart summarize the distribution of armed status. The analysis reveals that the majority of individuals in the dataset—78.09%—are armed, with only 21.91% classified as unarmed. This substantial difference underscores the predominance of armed individuals within the sample. The bar chart visually reinforces this finding, with the "Yes" category towering over the "No" category. This imbalance raises questions about the underlying reasons for such a distribution, whether cultural, legal, or related to other societal factors.

```{r armed-status-plot}
# Armed Status Distribution
armed_counts <- table(cleaned_data$armed)
armed_percentages <- prop.table(armed_counts) * 100
armed_summary <- data.frame(
  Armed_Status = names(armed_counts),
  Count = as.integer(armed_counts),
  Percentage = round(as.numeric(armed_percentages), 2) # Round percentages for clarity
)
pander(armed_summary, caption  = "Armed Status Distribution")


# Plot 1: Distribution of Armed Status
plot3 <- ggplot(cleaned_data, aes(x = armed)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of Armed Status", x = "Armed Status", y = "Count")
plot3
```

### Relationship Between Armed Status and Age

The relationship between age and armed status is further explored through a box plot, which compares the age distribution for armed and unarmed individuals. The plot illustrates that the age ranges and medians for both groups are remarkably similar, suggesting no significant difference in age distribution between those who are armed and those who are not. This visual evidence supports the regression findings, reinforcing the conclusion that age alone is not a strong determinant of armed status. Nonetheless, the lack of significant variability in the data suggests that other contextual factors beyond age may play a more crucial role in influencing armed status.

```{r armed-age-plot}
# Plot 2: Relationship between 'armed' and 'age'
plot4 <- ggplot(cleaned_data, aes(x = armed, y = age)) +
  geom_boxplot(fill = "lightblue", outlier.colour = "red", outlier.shape = 1) +
  geom_jitter(width = 0.2, alpha = 0.2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Relationship between Armed Status and Age",
       x = "Armed Status",
       y = "Age")
plot4

```

```{r age-plot}
ggplot(cleaned_data, aes(x = age, fill = armed)) +
  geom_histogram(position = "fill", bins = 20) +
  labs(title = "Likelihood of Being Armed by Age", x = "Age", y = "Proportion")
```

### Unemployment Rate and Likelihood of Being Armed

The role of unemployment rate is captured in the histogram, which displays the proportion of armed individuals across varying levels of unemployment. While the logistic regression did not find this relationship to be statistically significant, the visualizations hint at a potential pattern. Specifically, the proportion of armed individuals appears to increase slightly at higher unemployment rates. This observation suggests that unemployment could contribute indirectly to the likelihood of being armed, potentially reflecting economic pressures or broader social dynamics at play. The trend, while subtle, calls for more granular investigations to uncover the precise mechanisms underlying this association.

```{r urate-plot}
ggplot(cleaned_data, aes(x = urate, fill = armed)) +
  geom_histogram(position = "fill", bins = 20) +
  labs(title = "Likelihood of Being Armed by Unemployment Rate", x = "Unemployment Rate", y = "Proportion")
```

## Conclusion

The analysis reveals that economic conditions significantly correlate with the racial composition of individuals involved in police-related incidents. Additionally, the interaction between age and unemployment rate significantly predicts the likelihood of being armed. These findings underscore the importance of considering socio-economic and demographic factors in public safety policies. Limitations of this study include \[limitations\]. Future work could explore \[suggestions for future research\].