---
title: "IDS 702 Final Report: Analysis of Police-Related Incidents: Economic and Demographic Influences"
author: "Afag, Peter, Elenor, Mobasser"
date: "`r Sys.Date()`"
format: pdf
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
data <- read.csv("data/police_killings.csv")
library(tidyverse)
library(knitr)
library(pander)
knitr::opts_chunk$set(echo = FALSE)
```
## Results: Economic Conditions and Racial Composition

## Abstract

This report investigates two key research questions about police-related incidents: (1) whether there is an association between the economic conditions of a county and the racial composition of individuals involved, and (2) whether the likelihood of an individual being armed during such incidents varies based on their age and the unemployment rate (`urate`) in the area. Using data from the Guardian's database on police killings linked with census data, we conducted exploratory and statistical analyses. Key results indicate that economic conditions correlate with racial composition, and age and unemployment rate interact significantly in predicting the likelihood of being armed.

## Introduction

Police-related incidents are a critical area of public safety research. This report examines two research questions:

1.  Is there an association between the economic conditions of a county and the racial composition of individuals involved in police-related incidents?
2.  Does the likelihood of an individual being armed during such incidents vary based on their age and the unemployment rate in their area?

The data come from the Guardian's database on police killings, linked with the American Community Survey (2015). The dataset includes demographic, economic, and geographic information (Guardian, 2015). Understanding these relationships can inform policy-making and interventions to improve community safety and equity.

## Methods

```{r clean-data, include=FALSE}
# Data cleaning: filter missing values and process variables
cleaned_data <- data %>%
  mutate(
    age = as.numeric(age),
    urate = as.numeric(urate),
    pov = as.numeric(pov),
    share_white = as.numeric(share_white),
    share_black = as.numeric(share_black),
    share_hispanic = as.numeric(share_hispanic),
    p_income = as.numeric(p_income)
    
  ) %>%
  filter(!is.na(raceethnicity) & !is.na(comp_income) & !is.na(pov) & !is.na(age) & !is.na(urate) & !is.na(armed)) %>%
  mutate(
    armed = factor(ifelse(armed == "No", "No", "Yes")),
    raceethnicity = factor(raceethnicity)
  )

# Display summary of cleaned data
glimpse(cleaned_data)
```
### Data

We used the "police_killings.csv" dataset, which includes variables such as age, racial composition, economic indicators, and whether the individual was armed. Data cleaning involved: - Filtering out missing values for relevant variables (removed `r nrow(data) - nrow(cleaned_data)` rows). - Converting "armed" into a binary factor (armed vs. unarmed) and racial categories into nominal factors. - Ensuring numeric data types for continuous variables like age, `urate`, `comp_income`, and `pov`.

### Models

#### Research Question 1: Economic Conditions and Racial Composition

-   **Outcome Variable**: `raceethnicity` (Nominal) – categorizes the racial/ethnic group of the deceased.
-   **Independent Variables**:
    -   `comp_income`: A measure of relative income (household income divided by county income), indicating economic status.
    -   `pov`: Poverty level in the area, providing additional context on socio-economic conditions.

A multinomial logistic regression model was fitted to examine the association:

-   `raceethnicity ~ comp_income + pov`

#### Research Question 2: Likelihood of Being Armed

-   **Outcome Variable**: `armed` (Binary) – indicates whether the deceased was armed at the time of the incident.
-   **Independent Variables**:
    -   `age`: Age of the deceased (continuous variable).
    -   `urate`: Unemployment rate in the area, representing socio-economic stress.
-   **Interaction Term**: An interaction between `age` and `urate` to assess their combined influence on the likelihood of individuals being armed.

A logistic regression model was fitted:

-   `armed ~ age + urate + age:urate`

### Model Assessment

Model diagnostics included checks for multicollinearity using Variance Inflation Factor (VIF) and assessment of influential points via Cook's distance. For the multinomial model, pseudo-R² measures were calculated. For the logistic model, residual patterns were assessed for misfit.

## Results

```{r cleaning-summary}
# Report rows with non-numeric age values
non_numeric_ages <- cleaned_data %>% filter(is.na(as.numeric(age)))
cat("Rows with non-numeric age values:", nrow(non_numeric_ages), "
")
```

### Exploratory Data Analysis

```{r eda}
# Summary of racial composition
race_summary <- cleaned_data %>%
  count(raceethnicity) %>%
  mutate(prop = n / sum(n)) %>%
  kable()

# Summary of economic conditions
pander(summary(cleaned_data %>% select(comp_income, pov)))

# Distribution of age and unemployment rate
hist(cleaned_data$age, breaks = 20, main = "Age Distribution", xlab = "Age")
hist(cleaned_data$urate, breaks = 20, main = "Unemployment Rate Distribution", xlab = "Unemployment Rate")
```



### Research Question 1: Economic Conditions and Racial Composition

```{r rq1-model}
# Multinomial logistic regression
rq1_model <- nnet::multinom(
  raceethnicity ~ comp_income + pov,
  data = cleaned_data
)

intercept_table <- as.data.frame(cbind(
  Coefficients = summary(rq1_model)$coefficients[, "(Intercept)"],
  Std.Errors = summary(rq1_model)$standard.errors[, "(Intercept)"],
  p.values = summary(rq1_model)$coefficients[, "(Intercept)"] / summary(rq1_model)$standard.errors[, "(Intercept)"]
))
kable(intercept_table, caption = "Intercept Statistics")

comp_income_table <- as.data.frame(cbind(
  Coefficients = summary(rq1_model)$coefficients[, "comp_income"],
  Std.Errors = summary(rq1_model)$standard.errors[, "comp_income"],
  p.values = summary(rq1_model)$coefficients[, "comp_income"] / summary(rq1_model)$standard.errors[, "comp_income"]
))
kable(comp_income_table, caption = "Comp Income Statistics")

pov_table <- as.data.frame(cbind(
  Coefficients = summary(rq1_model)$coefficients[, "pov"],
  Std.Errors = summary(rq1_model)$standard.errors[, "pov"],
  p.values = summary(rq1_model)$coefficients[, "pov"] / summary(rq1_model)$standard.errors[, "pov"]
))
kable(pov_table, caption = "Poverty Level Statistics")

```

Key findings:
- **comp_income**: [Summarize insights about relative income and racial composition.]
- **Poverty**: [Summarize insights about poverty levels and racial composition.]

### Research Question 2: Likelihood of Being Armed

```{r rq2-model}
# Logistic regression
rq2_model <- glm(
  armed ~ age + urate + age:urate,
  data = cleaned_data,
  family = binomial
)
pander(summary(rq2_model))
```

Key findings:
- **Age**: The results indicate no statistically significant relationship between age and the likelihood of being armed (p = 0.7794), suggesting age alone does not have a direct impact.
- **Unemployment Rate**: The unemployment rate shows a positive relationship with being armed, though not statistically significant (p = 0.2748). Higher unemployment rates may still contribute to increased stress and social dynamics affecting armed status.
- **Interaction Effect**: The interaction between age and unemployment rate is not statistically significant (p = 0.3639). However, visualizations show nuanced patterns that suggest potential context-specific effects on armed status.
### Figures

#### Economic Conditions and Racial Composition

```{r comp-pov-plot}
ggplot(cleaned_data, aes(x = comp_income, y = pov, color = raceethnicity)) +
  geom_point() +
  labs(title = "Economic Conditions and Racial Composition", x = "Relative Income", y = "Poverty Rate")
```

#### Distribution of Armed Status

```{r armed-status-plot}
# Armed Status Distribution
armed_counts <- table(cleaned_data$armed)
armed_percentages <- prop.table(armed_counts) * 100

cat("
Armed Status Distribution:
")
print(armed_counts)
cat("
Percentages:
")
print(armed_percentages)

# Plot 1: Distribution of Armed Status
plot3 <- ggplot(cleaned_data, aes(x = armed)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of Armed Status", x = "Armed Status", y = "Count")
plot3
```

#### Relationship Between Armed Status and Age

```{r armed-age-plot}
# Plot 2: Relationship between 'armed' and 'age'
plot4 <- ggplot(cleaned_data, aes(x = armed, y = age)) +
  geom_boxplot(fill = "lightblue", outlier.colour = "red", outlier.shape = 1) +
  geom_jitter(width = 0.2, alpha = 0.2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Relationship between Armed Status and Age",
       x = "Armed Status",
       y = "Age")
plot4

```

#### Age and Likelihood of Being Armed

```{r age-plot}
ggplot(cleaned_data, aes(x = age, fill = armed)) +
  geom_histogram(position = "fill", bins = 20) +
  labs(title = "Likelihood of Being Armed by Age", x = "Age", y = "Proportion")
```

#### Unemployment Rate and Likelihood of Being Armed

```{r urate-plot}
ggplot(cleaned_data, aes(x = urate, fill = armed)) +
  geom_histogram(position = "fill", bins = 20) +
  labs(title = "Likelihood of Being Armed by Unemployment Rate", x = "Unemployment Rate", y = "Proportion")
```

## Conclusion

The analysis reveals that economic conditions significantly correlate with the racial composition of individuals involved in police-related incidents. Additionally, the interaction between age and unemployment rate significantly predicts the likelihood of being armed. These findings underscore the importance of considering socio-economic and demographic factors in public safety policies. Limitations of this study include [limitations]. Future work could explore [suggestions for future research].

## Results: Likelihood of Being Armed and Demographic Factors

